---
title: "R|R Data Quality Assurance"
output: html_notebook
---

This notebook was run on `r Sys.Date()`

## Configuration
```{r packages, warning=FALSE, message=FALSE}
library(aws.s3)
library(arrow)
library(duckdb)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```

##  Create Connections to Datasets
This code establishes a connection to the .parquet files in SAND_DS S3 as arrow datasets and then again as duckDB tables.
```{r datasets}
# arrow datasets of nodes
da <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/distinctauthor")
inst <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/linked_institutions/")
art <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/projected_article/")

# arrow datasets of edges
aff <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/AFFILIATION/")
contr <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/CONTRIBUTION")
topics <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/TOPIC_OF_INTEREST")
collab <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/COLLABORATOR")

# register duckdb tables
con <- dbConnect(duckdb::duckdb())
duckdb_register_arrow(con, "distinctAuthor", da)
duckdb_register_arrow(con, "institutions", inst)
duckdb_register_arrow(con, "articles", art)
duckdb_register_arrow(con, "affiliated_to", aff)
duckdb_register_arrow(con, "contribution", contr)
duckdb_register_arrow(con, "topics_of_interest", topics)
duckdb_register_arrow(con, "collaborators", collab)
```

## Metrics
### CM1
% of distinctAuthors with at least one `contribution` edge.
```{r}
count_da <- dbGetQuery(
  con,
  "Select count(distinct id)
  from distinctAuthor"
)

count_contr_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from contribution"
)

count_contr_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Ratio of contribution edges to occurrences
```{r}
count_occ <- dbGetQuery(
  con,
  "SELECT SUM(occurrences)
  FROM distinctAuthor"
)

count_contr <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM contribution"
)

count_contr$`count_star()` / count_occ$`sum(occurrences)`
```

### CM2
% of distinctAuthors with at least one affiliation edge
```{r}
count_aff_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from affiliated_to"
)

count_aff_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with no affiliation edge
```{r}
count_da$`count(DISTINCT id)` - count_aff_distinct$`count(DISTINCT source)`
```

### CM3
% of distinctAuthors with at least one TOPIC_OF_INTERST
```{r}
count_topics_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from topics_of_interest"
)

count_topics_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

### CM4
% of distinctAuthors with at least on collaborated_with edge
```{r}
count_collab_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  FROM collaborators"
)

count_collab_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with no collaborators
```{r}
count_da$`count(DISTINCT id)` - count_collab_distinct$`count(DISTINCT source)`
```

### CM5
% of distinctAuthors with forename >3 characters in length
```{r}
count_namelength_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct id)
  FROM distinctAuthor
  WHERE length(forename) > 2"
)

count_namelength_distinct$`count(DISTINCT id)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with givenname <=3 characters in length
```{r}
count_da$`count(DISTINCT id)` - count_namelength_distinct$`count(DISTINCT id)`
```

Examples:
```{r}
dbGetQuery(
  con,
  "SELECT id, forename, lastname
  FROM distinctAuthor
  WHERE length(forename) < 3
  LIMIT 1000"
)

```

### AR-CM1
% of articles with a DOI
```{r}
count_article <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM articles"
)

count_hasDOI <- dbGetQuery(
  con,
  "SELECT COUNT(*)
  FROM articles
  WHERE doi IS NOT NULL
  AND TRIM(doi) <> ''"
)

count_hasDOI$`count_star()` / count_article$`count_star()`
```

### AR-CM2
% of articles with a pub_date
```{r}
count_hasPubDate <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM articles
  WHERE pub_date IS NOT NULL"
)

count_hasPubDate$`count_star()` / count_article$`count_star()`
```

## Fivenum Summaries and Thresholds
### What is the distribution of occurrences?
```{r}
dbGetQuery(con,
           "SELECT MIN(da.occurrences) as min,
           QUANTILE_CONT(da.occurrences, 0.25) as p25,
           MEDIAN(da.occurrences) as med,
           QUANTILE_CONT(da.occurrences, 0.75) as p75,
           QUANTILE_CONT(da.occurrences, 0.9) as p90,
           QUANTILE_CONT(da.occurrences, 0.95) as p95,
           QUANTILE_CONT(da.occurrences, 0.99) as p99,
           QUANTILE_CONT(da.occurrences, 0.999) as p999,
           QUANTILE_CONT(da.occurrences, 0.9999) as p9999,
           MAX(da.occurrences) as max
           FROM distinctAuthor da")
```

### What is the distribution of the number of collaborators
```{r}
dbGetQuery(con,
           "SELECT MIN(da.collaborators_count) as min,
           QUANTILE_CONT(da.collaborators_count, 0.25) as p25,
           MEDIAN(da.collaborators_count) as med,
           QUANTILE_CONT(da.collaborators_count, 0.75) as p75,
           QUANTILE_CONT(da.collaborators_count, 0.9) as p90,
           QUANTILE_CONT(da.collaborators_count, 0.95) as p95,
           QUANTILE_CONT(da.collaborators_count, 0.99) as p99,
           QUANTILE_CONT(da.collaborators_count, 0.999) as p999,
           QUANTILE_CONT(da.collaborators_count, 0.9999) as p9999,
           MAX(da.collaborators_count) as max
           FROM distinctAuthor da")
```
### What is the distribution of affiliations per author?
```{r}
affialtions_to_author <- aff |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  collect()

affialtions_to_author |>
  summarise(percentiles = c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0),
            count_aff = quantile(count, c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0)))
```







# Experiments and Scratch
```{r, eval=FALSE}
contr |>
  filter(source == "r4r:01JMF02071PWPPHT022S30NJQW") |>
  collect()


da_json <- open_dataset("s3://com-copyright-agpipeline-sandds/data/output/phyeng/json/graph/nodes/r4r_projection_authors/",
                        format = "json")
duckdb_register_arrow(con, "da_json", da_json)

test_1 <- dbGetQuery(
  con,
  "SELECT *
  FROM da_json
  WHERE persistent_identifier = 'r4r:01JMF02071PWPPHT022S30NJQW'"
)


top_topic <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/TOP_TOPIC/")
top_topic |> count() |> collect()

topics |> count() |> collect()

field_of_study <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/FIELD_OF_STUDY/")
field_of_study |> count() |> collect()


articles <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/article/")
articles |> count() |> collect()
fugazi_2 <- articles |> head(10) |> collect()


au_docs <- open_dataset("s3://com-copyright-agpipeline-sandds/data/output/phyeng/json/graph/nodes/", format = "json")
duckdb_register_arrow(con, "author_doc", au_docs)
dbGetQuery(
  con,
  "SELECT *
  FROM author_doc
  WHERE persistent_identifier = 'r4r:01JMF02071PWPPHT022S30NJQW'"
)


fugazi_3 <- dbGetQuery(
  con,
  "SELECT da.id, count(c.id)
  FROM distinctAuthor da
  JOIN contribution c on da.id = c.source
  WHERE length(forename) > 2
  GROUP BY da.id
  "
)


aff |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  filter(count > 4) |>
  count() |>
  collect()
```

## Thresholds
## ORCID
```{r}
r4rid_orcid <- dbGetQuery(
  con,
  "SELECT id, occurrences, orcid
  FROM distinctAuthor
  WHERE LEN(orcid) > 1"
)

r4rid_orcid_unnest <- tibble(
  r4r_id = r4rid_orcid$id,
  occurrences = r4rid_orcid$occurrences,
  orcid = r4rid_orcid$orcid) |>
  unnest_longer(orcid) |>
  filter(orcid != "")

id_orc_occ <- r4rid_orcid_unnest |>
  group_by(r4r_id, occurrences) |>
  summarise(orc_count = n())

fivenum(id_orc_occ$orc_count)
quantile(id_orc_occ$orc_count, c(.9,.99,.999, .9999))
```
