---
title: "R|R Data Quality Assurance"
output: html_notebook
---

This notebook was run on `r Sys.Date()`

## Configuration
```{r packages, warning=FALSE, message=FALSE}
library(aws.s3)
library(arrow)
library(duckdb)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```

##  Create Connections to Datasets
This code establishes a connection to the .parquet files in SAND_DS S3 as arrow datasets and then again as duckDB tables.
```{r datasets}
# arrow datasets of nodes
da <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/distinctauthor")
inst <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/linked_institutions/")
art <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/projected_article/")

# arrow datasets of edges
aff <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/AFFILIATION/")
contr <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/CONTRIBUTION")
topics <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/TOPIC_OF_INTEREST")
collab <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/COLLABORATOR")

# register duckdb tables
con <- dbConnect(duckdb::duckdb())
duckdb_register_arrow(con, "distinctAuthor", da)
duckdb_register_arrow(con, "institutions", inst)
duckdb_register_arrow(con, "articles", art)
duckdb_register_arrow(con, "affiliated_to", aff)
duckdb_register_arrow(con, "contribution", contr)
duckdb_register_arrow(con, "topics_of_interest", topics)
duckdb_register_arrow(con, "collaborators", collab)
```

## Metrics
### CM1
% of distinctAuthors with at least on `contribution` edge.
```{r}
count_da <- dbGetQuery(
  con,
  "Select count(distinct id)
  from distinctAuthor"
)

count_contr_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from contribution"
)

count_contr_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Ratio of contribution edges to occurrences
```{r}
count_occ <- dbGetQuery(
  con,
  "SELECT SUM(occurrences)
  FROM distinctAuthor"
)

count_contr <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM contribution"
)

count_contr$`count_star()` / count_occ$`sum(occurrences)`
```

### CM2
% of distinctAuthors with at least one affiliation edge
```{r}
count_aff_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from affiliated_to"
)

count_aff_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with no affiliation edge
```{r}
count_da$`count(DISTINCT id)` - count_aff_distinct$`count(DISTINCT source)`
```

### CM3
% of distinctAuthors with at least one TOPIC_OF_INTERST
```{r}
count_topics_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  from topics_of_interest"
)

count_topics_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

### CM4
% of distinctAuthors with at least on collaborated_with edge
```{r}
count_collab_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct source)
  FROM collaborators"
)

count_collab_distinct$`count(DISTINCT source)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with no collaborators
```{r}
count_da$`count(DISTINCT id)` - count_collab_distinct$`count(DISTINCT source)`
```

### CM5
% of distinctAuthors with forename >3 characters in length
```{r}
count_namelength_distinct <- dbGetQuery(
  con,
  "SELECT count(distinct id)
  FROM distinctAuthor
  WHERE length(forename) > 2"
)

count_namelength_distinct$`count(DISTINCT id)` / count_da$`count(DISTINCT id)`
```

Count of distinctAuthors with givenname <=3 characters in length
```{r}
count_da$`count(DISTINCT id)` - count_namelength_distinct$`count(DISTINCT id)`
```

Examples:
```{r}
dbGetQuery(
  con,
  "SELECT id, forename, lastname
  FROM distinctAuthor
  WHERE length(forename) < 3
  LIMIT 1000"
)

```

### AR-CM1
% of articles with a DOI
```{r}
count_article <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM articles"
)

count_hasDOI <- dbGetQuery(
  con,
  "SELECT COUNT(*)
  FROM articles
  WHERE doi IS NOT NULL
  AND TRIM(doi) <> ''"
)

count_hasDOI$`count_star()` / count_article$`count_star()`
```

### AR-CM2
% of articles with a pub_date
```{r}
count_hasPubDate <- dbGetQuery(
  con,
  "SELECT count(*)
  FROM articles
  WHERE pub_date IS NOT NULL"
)

count_hasPubDate$`count_star()` / count_article$`count_star()`
```







# Experiments and Scratch
```{r, eval=FALSE}
contr |>
  filter(source == "r4r:01JMF02071PWPPHT022S30NJQW") |>
  collect()


da_json <- open_dataset("s3://com-copyright-agpipeline-sandds/data/output/phyeng/json/graph/nodes/r4r_projection_authors/",
                        format = "json")
duckdb_register_arrow(con, "da_json", da_json)

test_1 <- dbGetQuery(
  con,
  "SELECT *
  FROM da_json
  WHERE persistent_identifier = 'r4r:01JMF02071PWPPHT022S30NJQW'"
)


top_topic <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/TOP_TOPIC/")
top_topic |> count() |> collect()

topics |> count() |> collect()

field_of_study <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/FIELD_OF_STUDY/")
field_of_study |> count() |> collect()


articles <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/article/")
articles |> count() |> collect()
fugazi_2 <- articles |> head(10) |> collect()
```
