---
title: "Affiliated Institution Statistics"
output: html_notebook
---

This notebook was run on `r Sys.Date()`

## Configuration
```{r packages, warning=FALSE, message=FALSE}
library(aws.s3)
library(arrow)
library(duckdb)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```

# Data
This code establishes a connection to the .parquet files in SAND_DS S3 as arrow datasets.
```{r datasets}
# arrow datasets
aff_to_edge <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/edges/IM_AFFILIATED_TO")
aff_to_edge_gran <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/edges/IM_AFFILIATED_TO_GRANULAR")

inst <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/nodes/linked_institutions/")
inst_gran <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/nodes/linked_institutions_granular/")

aff_edge <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/edges/AFFILIATION/")
aff_edge_gran <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/pubmed/covid/graph/edges/AFFILIATION_GRANULAR/")
```

# Questions
### How many linked institutions are there?
Before
```{r}
inst |> count() |> collect()
```
After
```{r}
inst_gran |> count() |> collect()
```
### How many affiliation relationships are there?
Before
```{r}
aff_edge |> count() |> collect()
```
After
```{r}
aff_edge_gran |> count() |> collect()
```

### What is the count of distinct institutions with affiliations?
Before
```{r}
aff_edge |> distinct(target) |> count() |> collect()
```
Before
```{r}
aff_edge_gran |> distinct(target) |> count() |> collect()
```

### What are the institutions with the most affiliations?
Before
```{r}
aff_edge |>
  group_by(target) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(25) |>
  collect()
```
After
```{r}
aff_edge_gran |>
  group_by(target) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(25) |>
  collect()
```
### What is the distribution of affiliations per author?
Before
```{r}
affiliations_to_author <- aff_edge |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  collect()

affiliations_to_author |>
  summarise(percentiles = c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0),
            count_aff = quantile(count, c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0)))
```
After
```{r}
affiliations_to_author_gran <- aff_edge_gran |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  collect()

affiliations_to_author_gran |>
  summarise(percentiles = c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0),
            count_aff = quantile(count, c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0)))
```






















