---
title: "R|R Data Quality Queries"
output: html_notebook
---

This notebook was run on `r Sys.Date()`

## Configuration
```{r packages, warning=FALSE, message=FALSE}
library(aws.s3)
library(arrow)
library(duckdb)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
```


##  Create Connections to Datasets
This code establishes a connection to the .parquet files in SAND_DS S3 as arrow datasets and then again as duckDB tables.
```{r datasets}
# arrow datasets
da <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/distinctauthor")
au_docs <- open_dataset("s3://com-copyright-agpipeline-sandds/data/output/phyeng/json/graph/nodes/", format = "json")
inst <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/linked_institutions/")
art <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/projected_article/")
aff <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/AFFILIATION/")

# duckdb tables
con <- dbConnect(duckdb::duckdb())
duckdb_register_arrow(con, "distinctAuthor", da)
duckdb_register_arrow(con, "author_doc", au_docs)
duckdb_register_arrow(con, "institutions", inst)
duckdb_register_arrow(con, "articles", art)
duckdb_register_arrow(con, "affiliated_to", aff)
```


## Descriptive Statistics and Quality Checks
### What is the count of distinctAuthors?
```{r}
da |> count() |> collect()
# au |> count() |> collect()
au_docs |> count() |> collect()
```
### What is the count of articles?
```{r article_count}
art |> count() |> collect()
```

### What is the count of institutions?
```{r inst_count}
inst |> count() |> collect()
```

### What is the count of affiliations?
```{r aff_count}
aff |> count() |> collect()
```

### What is the distribution of occurrences?
```{r}
dbGetQuery(con,
           "SELECT MIN(da.occurrences) as min,
           QUANTILE_CONT(da.occurrences, 0.25) as p25,
           MEDIAN(da.occurrences) as med,
           QUANTILE_CONT(da.occurrences, 0.75) as p75,
           QUANTILE_CONT(da.occurrences, 0.9) as p90,
           QUANTILE_CONT(da.occurrences, 0.95) as p95,
           QUANTILE_CONT(da.occurrences, 0.99) as p99,
           QUANTILE_CONT(da.occurrences, 0.999) as p999,
           QUANTILE_CONT(da.occurrences, 0.9999) as p9999,
           MAX(da.occurrences) as max
           FROM distinctAuthor da")
```
### Which author has the most (max) occurrences?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor da WHERE da.occurrences = 298")
```
### Which authors have the most (99.99 percentile) occurrences?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE occurrences >= 158")
```
Note: look at Valeria V. Tynchenko at 182 occurrences. One article found online: https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=9464340. Also co-authored with Vadim Tynchenko.

### What is the distribution of the number of collaborators
```{r}
dbGetQuery(con,
           "SELECT MIN(da.collaborators_count) as min,
           QUANTILE_CONT(da.collaborators_count, 0.25) as p25,
           MEDIAN(da.collaborators_count) as med,
           QUANTILE_CONT(da.collaborators_count, 0.75) as p75,
           QUANTILE_CONT(da.collaborators_count, 0.9) as p90,
           QUANTILE_CONT(da.collaborators_count, 0.95) as p95,
           QUANTILE_CONT(da.collaborators_count, 0.99) as p99,
           QUANTILE_CONT(da.collaborators_count, 0.999) as p999,
           QUANTILE_CONT(da.collaborators_count, 0.9999) as p9999,
           MAX(da.collaborators_count) as max
           FROM distinctAuthor da")
```
### Which author has the most (MAX) collaborators?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE collaborators_count = 6848")
```
### Which authors have the most (99.99 percentile) collaborators?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE collaborators_count >= 4709")
```

### What is the relationship between collaborator count and occurrences?
Uses a random sample of 1000 authors
```{r collab_occ}
collab_occ <- dbGetQuery(
  con,
  "SELECT da.collaborators_count, da.occurrences
  FROM distinctAuthor da
  WHERE da.occurrences > 5
  ORDER BY RANDOM()
  LIMIT 1000"
)

# Fit linear model
model <- lm(collaborators_count ~ occurrences, data = collab_occ)
 
# Compute residuals
collab_occ$above_line <- residuals(model) > 0  # TRUE if above the regression line
 
# Scatterplot with differentiation
ggplot(collab_occ, aes(x = occurrences, y = collaborators_count, color = above_line)) +
  geom_point() +
  # geom_smooth(method = "lm") +
  scale_color_manual(values = c("skyblue", "orange")) +  # Choose colors as needed
  labs(color = "Above Regression Line")

# Histograph with differentiation
ggplot(collab_occ, aes(x = collaborators_count, fill = above_line)) +
  geom_histogram(binwidth = 50) +  
  scale_fill_manual(values = c("skyblue", "orange")) +  
  labs(x = "Collaborators Count", y = "Frequency", fill = "Above Regression Line")

```

### What is the distribution of the degree of belief
```{r}
dbGetQuery(con,
           "SELECT MIN(da.degree_of_belief) as min,
           QUANTILE_CONT(da.degree_of_belief, 0.25) as p25,
           MEDIAN(da.degree_of_belief) as med,
           QUANTILE_CONT(da.degree_of_belief, 0.75) as p75,
           QUANTILE_CONT(da.degree_of_belief, 0.9) as p90,
           QUANTILE_CONT(da.degree_of_belief, 0.95) as p95,
           QUANTILE_CONT(da.degree_of_belief, 0.99) as p99,
           QUANTILE_CONT(da.degree_of_belief, 0.999) as p999,
           QUANTILE_CONT(da.degree_of_belief, 0.9999) as p9999,
           MAX(da.degree_of_belief) as max
           FROM distinctAuthor da")
```

### What is the distribution of the degree of belief for authors with more than 1 article?
```{r}
dbGetQuery(con,
           "SELECT MIN(da.degree_of_belief) as min,
           QUANTILE_CONT(da.degree_of_belief, 0.25) as p25,
           MEDIAN(da.degree_of_belief) as med,
           QUANTILE_CONT(da.degree_of_belief, 0.75) as p75,
           QUANTILE_CONT(da.degree_of_belief, 0.9) as p90,
           QUANTILE_CONT(da.degree_of_belief, 0.95) as p95,
           QUANTILE_CONT(da.degree_of_belief, 0.99) as p99,
           QUANTILE_CONT(da.degree_of_belief, 0.999) as p999,
           QUANTILE_CONT(da.degree_of_belief, 0.9999) as p9999,
           MAX(da.degree_of_belief) as max
           FROM distinctAuthor da
           WHERE da.occurrences > 1")
```
### Scatterplot of DoB to Occurrences for Authors with > 1 Occurrences (10000 random)
```{r}
da_1000 <- dbGetQuery(con,
           "SELECT da.occurrences, da.degree_of_belief
           FROM distinctAuthor da
           WHERE da.occurrences > 1
           ORDER BY RANDOM()
           LIMIT 10000")

p1 <- ggplot(da_1000, aes(x=occurrences, y=degree_of_belief)) +
  geom_point() +
  geom_smooth(method = "lm")

p1
```

### How many distinctAuthors have a Degree of Belief of 1.0?
```{r}
dbGetQuery(con,
           "SELECT count(*)
           FROM distinctAuthor da
           WHERE da.degree_of_belief = 1")
```
### How many distinctAuthors have a Degree of Belief of 1.0 and only 1 article (cluster size = 1)?
```{r}
dbGetQuery(con,
           "SELECT count(*)
           FROM distinctAuthor da
           WHERE da.degree_of_belief = 1
           AND da.occurrences = 1")
```
### How many linked institutions are there? (repeat)
```{r}
inst |> count() |> collect()
```
### What are the properties of the AFFILIATION edge?
```{r}
aff |> head(10) |> collect()
```
### How many affiliation relationships are there?
```{r}
aff |> count() |> collect()
```
### What is the count of distinct institutions with affiliations?
```{r}
aff |> distinct(target) |> count() |> collect()
```
### What are the institutions with the most affiliations?
```{r}
aff |>
  group_by(target) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(25) |>
  collect()
```
### What is the distribution of affiliations per author?
```{r}
affialtions_to_author <- aff |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  collect()

affialtions_to_author |>
  summarise(percentiles = c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0),
            count_aff = quantile(count, c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0)))
```
### Plots of Affiliations - Occurrences

Note: this doesn't seen really helpful until we have more affiliations
```{r}
aff_occ <- dbGetQuery(
  con,
  "SELECT af.source, da.occurrences, count(*) as aff_count
  FROM distinctAuthor da
  LEFT JOIN affiliated_to af on af.source = da.id
  GROUP BY af.source, da.occurrences"
)

aff_occ_2 <- aff_occ |> filter(is.na(source) == FALSE)

# Fit linear model
model_aff <- lm(aff_count ~ occurrences, data = aff_occ_2)
 
# Compute residuals
aff_occ_2$above_line <- residuals(model_aff) > 0  # TRUE if above the regression line
 
# Scatterplot with differentiation
ggplot(aff_occ_2, aes(x = occurrences, y = aff_count, color = above_line)) +
  geom_point() +
  # geom_smooth(method = "lm") +
  scale_color_manual(values = c("skyblue", "orange")) +  # Choose colors as needed
  labs(color = "Above Regression Line")

# Histograph with differentiation
ggplot(aff_occ_2, aes(x = aff_count, fill = above_line)) +
  geom_histogram(binwidth = 2) +  
  scale_fill_manual(values = c("skyblue", "orange")) +  
  labs(x = "Affiliations Count", y = "Frequency", fill = "Above Regression Line")
```

### In which countries are the institutions with the most affiliations?
```{r}
institutions <- inst |> collect()
affiliations <- aff |> collect()

affiliations |>
  left_join(institutions, by = c("target" = "id")) |>
  group_by(country_name) |>
  summarise(count = n()) |>
  arrange(desc(count))

```
### Plot of Affiliations by Pub Year
```{r}
# data munging
counts_year <- unlist(affiliations$publication_info)
counts_year_df <- data.frame(
  pubyear_count = counts_year) |>
  separate(pubyear_count, into = c("pub_year", "count"), sep = ":")

# aggregate
count_year_summarise <- counts_year_df |>
  group_by(pub_year) |>
  summarise(total = sum(as.numeric(count)))

ggplot(count_year_summarise, aes(x = pub_year, y = total)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Affiliations by Publication Year")
```
## ORCID
```{r}
r4rid_orcid <- dbGetQuery(
  con,
  "SELECT id, occurrences, orcid
  FROM distinctAuthor
  WHERE LEN(orcid) > 1"
)

r4rid_orcid_unnest <- tibble(
  r4r_id = r4rid_orcid$id,
  occurrences = r4rid_orcid$occurrences,
  orcid = r4rid_orcid$orcid) |>
  unnest_longer(orcid) |>
  filter(orcid != "")

id_orc_occ <- r4rid_orcid_unnest |>
  group_by(r4r_id, occurrences) |>
  summarise(orc_count = n())

ggplot(id_orc_occ, aes(x=occurrences, y=orc_count)) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
# Fit linear model
model_orcid <- lm(orc_count ~ occurrences, data = id_orc_occ)
 
# Compute residuals
id_orc_occ$above_line <- residuals(model_orcid) > 0  # TRUE if above the regression line

temp <- id_orc_occ |> filter(above_line == FALSE)
fivenum(temp$orc_count)
quantile(temp$orc_count, c(.9,.95,.99,.999,1))

```
## Name Variants
```{r}
r4rid_namevar <- dbGetQuery(
  con,
  "SELECT id, occurrences, name_variants
  FROM distinctAuthor
  WHERE occurrences > 5"
)

r4rid_namevar_unnest <- tibble(
  r4r_id = r4rid_namevar$id,
  occurrences = r4rid_namevar$occurrences,
  namevar = r4rid_namevar$name_variants) |>
  unnest_longer(namevar)

id_nv_occ <- r4rid_namevar_unnest |>
  group_by(r4r_id, occurrences) |>
  summarise(namevar_count = n())

# set.seed(4567)
# id_nv_occ_sample <- id_nv_occ |>
#   mutate(random = runif(1)) |>
#   arrange(random) |>
#   head(10000)

ggplot(id_nv_occ, aes(x=occurrences, y=namevar_count)) +
  geom_point() +
  geom_smooth(method = "lm")

max(id_nv_occ$namevar_count)
```
### Name Variants, Occurrences, DoB
```{r}
r4rid_namevar_dob <- dbGetQuery(
  con,
  "SELECT id, occurrences, name_variants, degree_of_belief
  FROM distinctAuthor
  WHERE occurrences > 5"
)

r4rid_namevar_dob_unnest <- tibble(
  r4r_id = r4rid_namevar_dob$id,
  occurrences = r4rid_namevar_dob$occurrences,
  dob = r4rid_namevar_dob$degree_of_belief,
  namevar = r4rid_namevar_dob$name_variants) |>
  unnest_longer(namevar)

id_nv_occ_dob <- r4rid_namevar_dob_unnest |>
  group_by(r4r_id, occurrences, dob) |>
  summarise(namevar_count = n())

plot_ly(data = id_nv_occ_dob, x=~occurrences, y=~namevar_count, z=~dob,
        type="scatter3d", mode="markers",
        jitter = 0)

# set.seed(4567)
# id_nv_occ_sample <- id_nv_occ |>
#   mutate(random = runif(1)) |>
#   arrange(random) |>
#   head(10000)

# ggplot(id_nv_occ, aes(x=occurrences, y=namevar_count)) +
#   geom_point() +
#   geom_smooth(method = "lm")
# 
# max(id_nv_occ$namevar_count)

```

## END OF DOC








