---
title: "R|R Data Quality Queries"
output: html_notebook
---

## Configuration
```{r}
library(aws.s3)
library(arrow)
library(duckdb)
library(dplyr)
library(ggplot2)
library(tidyr)
```


##  Create Connections to Datasets
```{r}
# arrow datasets
da <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/distinctauthor")
# au <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/author/")
au_docs <- open_dataset("s3://com-copyright-agpipeline-sandds/data/output/phyeng/json/graph/nodes/", format = "json")
inst <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/linked_institutions/")
art <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/nodes/projected_article/")
aff <- open_dataset("s3://com-copyright-agpipeline-sandds/data/internal/parquet/r4r/phyeng/graph/edges/AFFILIATION/")

# duckdb tables
con <- dbConnect(duckdb::duckdb())
duckdb_register_arrow(con, "distinctAuthor", da)
# duckdb_register_arrow(con, "author", au)
duckdb_register_arrow(con, "author_doc", au_docs)
duckdb_register_arrow(con, "institutions", inst)
duckdb_register_arrow(con, "articles", art)
duckdb_register_arrow(con, "affiliated_to", aff)
```


## Descriptive Statistics and Quality Checks
### What is the count of distinctAuthors?
```{r}
da |> count() |> collect()
# au |> count() |> collect()
au_docs |> count() |> collect()
```
### What is the distribution of occurrences?
```{r}
dbGetQuery(con,
           "SELECT MIN(da.occurrences) as min,
           QUANTILE_CONT(da.occurrences, 0.25) as p25,
           MEDIAN(da.occurrences) as med,
           QUANTILE_CONT(da.occurrences, 0.75) as p75,
           QUANTILE_CONT(da.occurrences, 0.9) as p90,
           QUANTILE_CONT(da.occurrences, 0.95) as p95,
           QUANTILE_CONT(da.occurrences, 0.99) as p99,
           QUANTILE_CONT(da.occurrences, 0.999) as p999,
           QUANTILE_CONT(da.occurrences, 0.9999) as p9999,
           MAX(da.occurrences) as max
           FROM distinctAuthor da")
```
### Which author has the most (max) occurrences?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE occurrences = 516")
```
### Which authors have the most (99.99 percentile) occurrences?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE occurrences >= 158")
```
Note: look at Valeria V. Tynchenko at 182 occurrences. One article found online: https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=9464340. Also co-authored with Vadim Tynchenko.

### What is the distribution of the number of collaborators
```{r}
dbGetQuery(con,
           "SELECT MIN(da.collaborators_count) as min,
           QUANTILE_CONT(da.collaborators_count, 0.25) as p25,
           MEDIAN(da.collaborators_count) as med,
           QUANTILE_CONT(da.collaborators_count, 0.75) as p75,
           QUANTILE_CONT(da.collaborators_count, 0.9) as p90,
           QUANTILE_CONT(da.collaborators_count, 0.95) as p95,
           QUANTILE_CONT(da.collaborators_count, 0.99) as p99,
           QUANTILE_CONT(da.collaborators_count, 0.999) as p999,
           QUANTILE_CONT(da.collaborators_count, 0.9999) as p9999,
           MAX(da.collaborators_count) as max
           FROM distinctAuthor da")
```
### Which author has the most (MAX) collaborators?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE collaborators_count = 7384")
```
### Which authors have the most (99.99 percentile) collaborators?
```{r}
dbGetQuery(con,
           "SELECT * FROM distinctAuthor WHERE collaborators_count >= 4342")
```
### What is the distribution of the degree of belief
```{r}
dbGetQuery(con,
           "SELECT MIN(da.degree_of_belief) as min,
           QUANTILE_CONT(da.degree_of_belief, 0.25) as p25,
           MEDIAN(da.degree_of_belief) as med,
           QUANTILE_CONT(da.degree_of_belief, 0.75) as p75,
           QUANTILE_CONT(da.degree_of_belief, 0.9) as p90,
           QUANTILE_CONT(da.degree_of_belief, 0.95) as p95,
           QUANTILE_CONT(da.degree_of_belief, 0.99) as p99,
           QUANTILE_CONT(da.degree_of_belief, 0.999) as p999,
           QUANTILE_CONT(da.degree_of_belief, 0.9999) as p9999,
           MAX(da.degree_of_belief) as max
           FROM distinctAuthor da")
```

### What is the distribution of the degree of belief for authors with more than 1 article?
```{r}
dbGetQuery(con,
           "SELECT MIN(da.degree_of_belief) as min,
           QUANTILE_CONT(da.degree_of_belief, 0.25) as p25,
           MEDIAN(da.degree_of_belief) as med,
           QUANTILE_CONT(da.degree_of_belief, 0.75) as p75,
           QUANTILE_CONT(da.degree_of_belief, 0.9) as p90,
           QUANTILE_CONT(da.degree_of_belief, 0.95) as p95,
           QUANTILE_CONT(da.degree_of_belief, 0.99) as p99,
           QUANTILE_CONT(da.degree_of_belief, 0.999) as p999,
           QUANTILE_CONT(da.degree_of_belief, 0.9999) as p9999,
           MAX(da.degree_of_belief) as max
           FROM distinctAuthor da
           WHERE da.occurrences > 1")
```
### Scatterplot of DoB to Occurrences for Authors with > 1 Occurrences (10000 random)
```{r}
da_1000 <- dbGetQuery(con,
           "SELECT da.occurrences, da.degree_of_belief
           FROM distinctAuthor da
           WHERE da.occurrences > 1
           ORDER BY RANDOM()
           LIMIT 10000")

p1 <- ggplot(da_1000, aes(x=occurrences, y=degree_of_belief)) +
  geom_point() +
  geom_smooth(method = "lm")

p1
```

### How many distinctAuthors have a Degree of Belief of 1.0?
```{r}
dbGetQuery(con,
           "SELECT count(*)
           FROM distinctAuthor da
           WHERE da.degree_of_belief = 1")
```
### How many distinctAuthors have a Degree of Belief of 1.0 and only 1 article (cluster size = 1)?
```{r}
dbGetQuery(con,
           "SELECT count(*)
           FROM distinctAuthor da
           WHERE da.degree_of_belief = 1
           AND da.occurrences = 1")
```
### How many linked institutions are there?
```{r}
inst |> count() |> collect()
```
### What are the properties of the AFFILIATION edge?
```{r}
aff |> head(10) |> collect()
```
### How many affiliation relationships are there?
```{r}
aff |> count() |> collect()
```
### What is the count of distinct institutions with affiliations?
```{r}
aff |> distinct(target) |> count() |> collect()
```
### What are the institutions with the most affiliations?
```{r}
aff |>
  group_by(target) |>
  summarise(count = n()) |>
  arrange(desc(count)) |>
  head(25) |>
  collect()


```
### What is the distribution of affiliations per author?
```{r}
affialtions_to_author <- aff |>
  group_by(source) |>
  summarise(count = n(),
            ) |>
  collect()

affialtions_to_author |>
  summarise(percentiles = c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0),
            count_aff = quantile(count, c(0,.25,0.5,0.75,0.9,0.99,0.999,0.9999, 1.0)))
```
# Plot of Affiliations by Pub Year
```{r}
# data munging
counts_year <- unlist(affiliations$publication_info) 
counts_year_df <- data.frame(
  pubyear_count = counts_year) |>
  separate(pubyear_count, into = c("pub_year", "count", sep = ":"))


```

# In which countries are the institutions with the most affiliations?
```{r}
institutions <- inst |> collect()
affiliations <- aff |> collect()

affiliations |>
  left_join(institutions, by = c("target" = "id")) |>
  group_by(country_name) |>
  summarise(count = n()) |>
  arrange(desc(count))

```

## Next Question


## END OF DOC
